---
title: "fecal-mifish-mockComm-contamination"
author: "diana baetscher"
date: "2026-05-15"
output: html_document
---

Katie Luxa noticed that there were species included in the decontaminated metabarcoding data output that realistically, could not be found in the Bering Sea near Bogoslof. Upon closer examination, many of the species are ones that were part of the mock communities used for Lia's nearshore eelgrass paper around the same time.

Digging deeper, it turns out that the mock communities were sequenced before the fecal samples and nothing else was sequenced in between. I'm guessing this is a bigger issue because these are mock communities generated with gBlocks and maybe are more resilient than amplicons generated from samples.

Given this discovery, I'm trying to figure out the best way to filter reads for species from the mock communities without removing all of the reads for species that are truly part of the fur seal diets.


I think my best bet is looking at the positive and negative controls and the ASVs and species identities. 
Then, to pair this analysis to the previous 'fecal-mifish-decontamination.Rmd'.


```{r load-libraries}
library(tidyverse)

collapsed_taxonomy <- read_csv("csv_outputs/inrange_collapsed_fecal_mifish_taxonomy.csv")

collapsed_taxonomy %>%
  filter(str_detect(taxon, "Oncorhynchus"))

```

```{r load-sample-info}
sample_metadata <- readxl::read_xlsx("../data/NFS_metadata_20241001.xlsx")

#illumina output changed "_" to "-"
# sample_metadata$sample_ID <- gsub("_", "-", sample_metadata$sample_ID) 
# sample_metadata$sample_ID_date <- gsub("_", "-", sample_metadata$sample_ID_date) 
```

check sequence table outputs
```{r}
asv_table <- read.csv("csv_outputs/NFSfecal_MiFish_ASVtable.csv") %>%
  rename(sample = X)

# reformat
sample_df <- asv_table %>%
  pivot_longer(2:length(asv_table), names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  separate(sample, into = c("ablg", "plate", "replicate"), remove = F)

head(sample_df)
```

# Look at positive and negative controls for mock community species

```{r positive-control}
sample_df %>%
  filter(ablg == "PC") %>%
  left_join(., collapsed_taxonomy) %>%
  ggplot(aes(x = sample, y = reads, fill = taxon)) +
  geom_bar(stat = "identity")


```
The contamination in the PC look reasonable except for the Sebastes and Cymatogaster aggregata. Hard to say on the Gadus. But, I could safely remove that number of reads of each of those species from all samples...


```{r negative-controls}
sample_df %>%
  filter(ablg == "NC") %>%
  left_join(., collapsed_taxonomy) %>%
  filter(!taxon %in% c("Callorhinus ursinus", "Leuroglossus schmidti")) %>%
  ggplot(aes(x = sample, y = reads, fill = taxon)) +
  geom_bar(stat = "identity") 

```
```{r}
mocks <- readxl::read_xlsx("../data/mock_community_species_contamination.xlsx", sheet = "mock_communities")


```

```{r}
sample_df %>%
  filter(ablg == "NC") %>%
  left_join(., collapsed_taxonomy) %>%
  inner_join(., mocks, by = c("taxon" = "Scientific name")) %>%
   ggplot(aes(x = sample, y = reads, fill = taxon)) +
  geom_bar(stat = "identity") 
```
It seems like these are all likely candidates for removal from scat samples. 

```{r}
sample_df %>%
  left_join(., collapsed_taxonomy) %>%
  inner_join(., mocks, by = c("taxon" = "Scientific name")) %>%
  filter(!is.na(taxon)) %>%
  filter(!str_detect(taxon, "Oncorhynchus")) %>%
  ggplot(aes(x = sample, y = reads, fill = taxon)) +
  geom_bar(stat = "identity")  +
  theme(
    axis.text.x = element_text(angle = 90, hjust = .95, size = 7)
  ) +
  facet_grid(cols = vars(plate), scales = "free", space = "free")
  


ggsave("pdf_outputs/mockCommunityContamination.pdf", width = 12, height = 5)
```

Looking at the plate effect - clearly, plate2 has a lot more contamination from the mock community species. I wonder if I drop that replicate and then filter out the small number of reads for species that are almost certainly contamination in the other plates, how clean will the data be?

Go through the previous filtering steps from the decontamination Rmd, but then drop the second plate.

# account for likely contaminants 

- tag-jumping
- negative PCR controls
- field negatives


## Step 1. Account for tag-jumping by using the positive controls 

subtract the proportion of reads that jumped into the positive control samples from each environmental sample 

```{r tag-jumping-by-ASVs-in-PC}
# on a per-plate basis, what % of each ASV should be subtracted from the samples?
asv_prop_df <- sample_df %>%
  group_by(sample) %>%
  mutate(total_reads = sum(reads)) %>%
  mutate(asv_prop = reads/total_reads)
  
```

```{r which-asv-proportions-to-remove}
# based on asvs in the PC
asv_prop_to_remove <- asv_prop_df %>%
  filter(ablg == "PC") %>%
  ungroup() %>%
  select(plate, ASV, asv_prop) %>%
  rename(prop_to_subtract = asv_prop)

```


```{r remove-asvs-props-from-PC-in-samples}
pc_filtered_sample_df <- asv_prop_df %>%
  left_join(., asv_prop_to_remove) %>%
  mutate(new_asv_prop = ifelse(!is.na(prop_to_subtract), asv_prop-prop_to_subtract, asv_prop)) %>%
  mutate(new_asv_prop = ifelse(new_asv_prop < 0, 0, new_asv_prop)) %>% # remove negative numbers
  #filter(new_asv_prop > 0) %>%
  mutate(reads_removed = prop_to_subtract*total_reads) %>%
  select(-prop_to_subtract, -asv_prop)

pc_filtered_sample_df
```

## Step 2. Account for contaminants in positive and negative controls 

next we will remove ASVs that only occur in controls and not in environmental samples. 

number of reads
```{r}
reads_per_type_ASV <- pc_filtered_sample_df %>%
  left_join(., sample_metadata, by = c("ablg" = "extraction_ID")) %>% 
  mutate(sample_type = ifelse(is.na(sample_type) & ablg == "NC", "pcr_blank", sample_type)) %>%
  mutate(sample_type = ifelse(is.na(sample_type) & ablg == "PC", "positive_control", sample_type)) %>%
  group_by(ASV, sample_type) %>%
  summarize(TotalReadsPerASV = sum(reads, na.rm = TRUE)) %>%
  arrange(ASV)

```

what ASVs have no reads in samples, but reads in the controls? 
```{r}
not_in_samples <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(is.na(sample)) %>% # here, Kim had a numeric, but mine are NA
  replace(is.na(.), 0)

not_in_samples
```

what ASVs do have reads in samples, but more reads in the controls? 
```{r}
more_in_pcr_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(pcr_blank > sample)
head(more_in_pcr_blanks)

more_in_extraction_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(extraction_blank > sample)
head(more_in_extraction_blanks)

more_in_pc_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(positive_control > sample)
head(more_in_pc_blanks)

more_in_fb_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(field_blank > sample)
head(more_in_fb_blanks)
```

remove these from the asv table
```{r}
asv_table_filter2 <- pc_filtered_sample_df %>%
  filter(!ASV %in% not_in_samples$ASV) %>%
  select(-reads_removed, -total_reads, -new_asv_prop)
 # filter(!ASV %in% more_in_pcr_blanks$ASV) %>%
  #filter(!ASV %in% more_in_extraction_blanks$ASV) %>%
  #filter(!ASV %in% more_in_fb_blanks$ASV)

# Let's stick with ASVs that are just in the controls for filtering
```


## Step 3. Remove ASVs without taxonomic ID 

now lets see how many of these ASVs have taxonomic IDs (these are not final tax ids)
```{r output-from-blastn-and-taxonkit}
taxonomy <- read_csv("csv_outputs/inrange_collapsed_fecal_mifish_taxonomy.csv")

combined_df <- asv_table_filter2 %>%
  left_join(., taxonomy) 
```


```{r remove-ASVs-without-taxonomy}
combined_df_noNA <- combined_df %>%
  filter(!is.na(taxon)) 

combined_df_noNA %>%
  group_by(taxon) %>%
  summarise(sum(reads))

```

```{r filter-controls-analyze-samples}
clean_df_filtered <- combined_df_noNA %>%
  left_join(., sample_metadata, by = c("ablg" = "extraction_ID")) %>%
  filter(plate != "plate2") %>%
  filter(sample_type == "sample") %>%
  filter(!taxon %in% c("Acipenser fulvescens", "Acipenser", "Callorhinus ursinus", "Eumetopias jubatus")) %>%
  group_by(sample, taxon) %>%
  mutate(taxon_reads = sum(reads)) %>%
  select(-ASV, -reads) %>%
  unique() %>%
  ungroup() %>%
  group_by(sample) %>%
  mutate(total_reads = sum(taxon_reads)) %>%
  select(sample, ablg, plate, replicate, taxon_reads, taxon, total_reads, everything()) %>%
  mutate(prop_taxon = taxon_reads/total_reads)
  

clean_df_filtered %>%
  ggplot(aes(x = ablg, y = prop_taxon, fill = taxon)) +
  geom_bar(stat = "identity") +
  facet_grid(cols = vars(replicate), rows = vars(plate)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90)
  ) 

```

```{r check-on-contamination-with-plate2-removed}
clean_df_filtered %>%
  group_by(taxon) %>%
  summarise(sum(taxon_reads)) %>%
  ggplot(aes(x = reorder(taxon, `sum(taxon_reads)`), y = `sum(taxon_reads)`)) +
  geom_bar(stat = "identity") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95)
  )

```

From this list, we definitely can remove the mock community species that are clearly wrong.

```{r}
# remove obvious species and then check replicates for Sebastes and Gadus
clean_df <- clean_df_filtered %>%
  filter(!taxon %in% c("Ammodytes personatus", "Cymatogaster aggregata", "Lepidopsetta", "Leptocottus armatus", "Lumpenus sagitta", "Ophiodon elongatus", "Pholis laeta", "Syngnathus leptorhynchus", "Sebastes", "Ammodytes hexapterus")) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(taxon_reads)) %>%
  select(sample, ablg, plate, replicate, taxon_reads, taxon, total_reads, everything()) %>%
  mutate(prop_taxon = taxon_reads/total_reads) 

clean_df %>%
  ggplot(aes(x = ablg, y = prop_taxon, fill = reorder(taxon, -prop_taxon))) +
  geom_bar(stat = "identity") +
  facet_grid(rows = vars(plate), cols = vars(replicate), space = "free", scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.position = "bottom"
  ) +
  scale_y_continuous(expand = c(0,0), breaks = c(0.25, 0.75)) +
  labs(fill = "",
       y = "read proportion",
       x = "sample")
  

ggsave("pdf_outputs/decon_fecal_data.pdf", width = 10, height = 8)
```
The presence of Sebastes in low read counts across so many samples seems dubious. For these diet samples, each diet has relatively few constituents.

CGATGT
GGATAA

Can I summarize all of the taxa and reads that were removed?
```{r}
working_df <- combined_df_noNA %>%
  left_join(., sample_metadata, by = c("ablg" = "extraction_ID")) %>%
  filter(plate != "plate2") %>%
  filter(sample_type == "sample") %>%
  filter(!taxon %in% c("Acipenser fulvescens", "Acipenser", "Callorhinus ursinus", "Eumetopias jubatus")) %>%
   filter(!taxon %in% c("Ammodytes personatus", "Cymatogaster aggregata", "Lepidopsetta", "Leptocottus armatus", "Lumpenus sagitta", "Ophiodon elongatus", "Pholis laeta", "Syngnathus leptorhynchus", "Sebastes", "Ammodytes hexapterus"))

sample_df %>%
  left_join(., taxonomy) %>%
  filter(!is.na(taxon)) %>%
  anti_join(., working_df) %>%
  group_by(taxon) %>%
  summarise(taxon_reads = sum(reads)) 
```




