---
title: "fecal-mifish-decontamination"
author: "diana baetscher"
date: "2024-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Borrowing from Kim's Bering Sea data decon procedure...


```{r load-libraries}
library(tidyverse)

```


```{r load-sample-info}
sample_metadata <- readxl::read_xlsx("../data/NFS_metadata_20241001.xlsx")

#illumina output changed "_" to "-"
# sample_metadata$sample_ID <- gsub("_", "-", sample_metadata$sample_ID) 
# sample_metadata$sample_ID_date <- gsub("_", "-", sample_metadata$sample_ID_date) 
```

check sequence table outputs
```{r}
asv_table <- read.csv("csv_outputs/NFSfecal_MiFish_ASVtable.csv") %>%
  rename(sample = X)

# reformat
sample_df <- asv_table %>%
  pivot_longer(2:length(asv_table), names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  separate(sample, into = c("ablg", "plate", "replicate"), remove = F)

head(sample_df)
```


# account for likely contaminants 

- tag-jumping
- negative PCR controls
- field negatives


## Step 1. Account for tag-jumping by using the positive controls 

subtract the proportion of reads that jumped into the positive control samples from each environmental sample 

```{r tag-jumping-by-ASVs-in-PC}
# on a per-plate basis, what % of each ASV should be subtracted from the samples?
asv_prop_df <- sample_df %>%
  group_by(sample) %>%
  mutate(total_reads = sum(reads)) %>%
  mutate(asv_prop = reads/total_reads)
  
asv_prop_df %>%
  filter(ablg == "PC") %>%
  ggplot(aes(x = sample, y = asv_prop, fill = ASV)) +
  geom_bar(stat = "identity") +
  facet_grid(row = vars(replicate)) +
  theme_minimal()

```
```{r which-asv-proportions-to-remove}
# based on asvs in the PC
asv_prop_to_remove <- asv_prop_df %>%
  filter(ablg == "PC") %>%
  ungroup() %>%
  select(plate, ASV, asv_prop) %>%
  rename(prop_to_subtract = asv_prop)

```


```{r remove-asvs-props-from-PC-in-samples}
pc_filtered_sample_df <- asv_prop_df %>%
  left_join(., asv_prop_to_remove) %>%
  mutate(new_asv_prop = ifelse(!is.na(prop_to_subtract), asv_prop-prop_to_subtract, asv_prop)) %>%
  mutate(new_asv_prop = ifelse(new_asv_prop < 0, 0, new_asv_prop)) %>% # remove negative numbers
  #filter(new_asv_prop > 0) %>%
  mutate(reads_removed = prop_to_subtract*total_reads) %>%
  select(-prop_to_subtract, -asv_prop)

pc_filtered_sample_df
```

Reads removed by tag-jumping:
```{r summarize-reads-removed}
pc_filtered_sample_df %>%
  filter(!is.na(reads_removed))
```





this is a summary of the number of reads removed by ASV and sample_ID
```{r summarize}
decontaminated_1 <- pc_filtered_sample_df %>%
  dplyr::select(sample, ASV, reads_removed) %>%
  filter(!is.na(replicate) &
           !is.na(reads_removed)) %>%
  group_by(ASV) %>%
  summarise(mean_reads = mean(reads_removed),
            reads_q.05 = quantile(reads_removed, probs=0.05),
            median_q.5 = median(reads_removed),
            reads_q.95 = quantile(reads_removed, probs=0.95)) %>%
  filter(mean_reads > 0) %>%
  filter(ASV != "ASV16") # remove sturgeon

decontaminated_1  

```


## Step 2. Account for contaminants in positive and negative controls 

next we will remove ASVs that only occur in controls and not in environmental samples. 

number of reads
```{r}
reads_per_type_ASV <- pc_filtered_sample_df %>%
  left_join(., sample_metadata, by = c("ablg" = "extraction_ID")) %>% 
  mutate(sample_type = ifelse(is.na(sample_type) & ablg == "NC", "pcr_blank", sample_type)) %>%
  mutate(sample_type = ifelse(is.na(sample_type) & ablg == "PC", "positive_control", sample_type)) %>%
  group_by(ASV, sample_type) %>%
  summarize(TotalReadsPerASV = sum(reads, na.rm = TRUE)) %>%
  arrange(ASV)

```

what ASVs have no reads in samples, but reads in the controls? 
```{r}
not_in_samples <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(is.na(sample)) %>% # here, Kim had a numeric, but mine are NA
  replace(is.na(.), 0)

not_in_samples
```


what ASVs do have reads in samples, but more reads in the controls? 
```{r}
more_in_pcr_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(pcr_blank > sample)
head(more_in_pcr_blanks)

more_in_extraction_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(extraction_blank > sample)
head(more_in_extraction_blanks)

more_in_pc_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(positive_control > sample)
head(more_in_pc_blanks)

more_in_fb_blanks <- reads_per_type_ASV %>%
  pivot_wider(names_from = "sample_type", values_from = c("TotalReadsPerASV")) %>%
  filter(sample > 1) %>%
  filter(field_blank > sample)
head(more_in_fb_blanks)
```

remove these from the asv table
```{r}
asv_table_filter2 <- pc_filtered_sample_df %>%
  filter(!ASV %in% not_in_samples$ASV) %>%
  select(-reads_removed, -total_reads, -new_asv_prop)
 # filter(!ASV %in% more_in_pcr_blanks$ASV) %>%
  #filter(!ASV %in% more_in_extraction_blanks$ASV) %>%
  #filter(!ASV %in% more_in_fb_blanks$ASV)

# Let's stick with ASVs that are just in the controls for filtering
```


## Step 3. Remove ASVs without taxonomic ID 

now lets see how many of these ASVs have taxonomic IDs (these are not final tax ids)
```{r output-from-blastn-and-taxonkit}
taxonomy <- read_csv("csv_outputs/fecal_taxonomy_extra_gadidSalmon.csv")

combined_df <- asv_table_filter2 %>%
  left_join(., taxonomy) 
```



```{r merge-data-frames-remove-NAs}
combined_df %>%
  filter(is.na(taxon)) %>%
  ungroup() %>%
  select(ASV) %>%
  unique()
```
166 ASVs without taxonomy (removed).

Some of those are selectively removed because they're off-target (i.e., primates, bacteria, etc.)

```{r remove-ASVs-without-taxonomy}
combined_df_noNA <- combined_df %>%
  filter(!is.na(taxon)) 

combined_df_noNA %>%
  group_by(taxon) %>%
  summarise(sum(reads))

```


```{r filter-controls-analyze-samples}

clean_df_filtered <- combined_df_noNA %>%
  left_join(., sample_metadata, by = c("ablg" = "extraction_ID")) %>%
  filter(sample_type == "sample") %>%
  filter(!taxon %in% c("Acipenser fulvescens", "Acipenser", "Callorhinus ursinus", "Eumetopias jubatus")) %>%
  group_by(sample, taxon) %>%
  mutate(taxon_reads = sum(reads)) %>%
  select(-ASV, -reads) %>%
  unique() %>%
  ungroup() %>%
  group_by(sample) %>%
  mutate(total_reads = sum(taxon_reads)) %>%
  select(sample, ablg, plate, replicate, taxon_reads, taxon, total_reads, everything()) %>%
  mutate(prop_taxon = taxon_reads/total_reads)
  

clean_df_filtered %>%
  ggplot(aes(x = ablg, y = prop_taxon, fill = taxon)) +
  geom_bar(stat = "identity") +
  facet_grid(cols = vars(replicate), rows = vars(plate)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90)
  ) 

```



### read in data from hard parts
(from Katie Luxa and Tonya Zeppelin at MML)


```{r hard-parts-data}
hardparts <- readxl::read_xlsx("../data/2023 NFS Bogoslof hard parts.xlsx")

# just keep the fishes 
# i.e., remove squids
hardparts_df <- hardparts %>%
  # select(ScientificName) %>%
  # unique() %>%
  filter(!ScientificName %in% c("Gonatidae", "Gonatopsis borealis/Berryteuthis magister", "Gonatus sp.", "NO PREY REMAINS", "Decapodiformes")) %>%
  rename(hardparts = ScientificName) %>%
  mutate(hardparts = ifelse(hardparts == "Northern lampfish", "Stenobrachius leucopsaurus", hardparts))

```
9 species of fishes.


```{r}
clean_df_filtered %>%
  #left_join(., hardparts_df, by = c("alternative_ID" = "DNASample")) %>%
  anti_join(., hardparts_df, by = c("alternative_ID" = "DNASample")) %>%
  filter(source != "fecal loop") %>%
  select(alternative_ID) %>%
  unique()
  
  
  # filter(!is.na(hardparts)) %>%
  # ggplot(aes(x = alternative_ID, y = prop_taxon, fill = taxon)) +
  # geom_bar(stat = "identity") +
  # facet_grid(cols = vars(replicate), rows = vars(plate)) +
  # theme_minimal() +
  # theme(
  #   legend.position = "none",
  #   axis.text.x = element_text(angle = 90)
  # ) 

```
From the hard parts, apparently we're just getting detections (y/n) for each of the species?

```{r}
hardparts_df %>%
  ggplot(aes(x = hardparts, fill = hardparts)) +
  geom_bar(stat = "count") +
  #facet_grid(cols = vars(replicate), rows = vars(plate)) +
  theme_minimal() +
  theme(
    #legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) 
  
```
Could we do presence/absence linear regression, hardparts vs. metabarcoding by species detections in the dataset?


Apply Bray-Curtis to metabarcoding replicates, then, assuming everything is similar, use the P/A data for comparison with hardparts.
```{r}
library(vegan)

# change df format for Bray-Curtis analysis
clean_df_filtered

```

```{r}
clean_df_filtered %>%
  arrange(taxon_reads) %>%
  # retaining taxa with 1 read seems crazy.
  ggplot(aes(taxon_reads)) +
  geom_histogram()
```
```{r}
min_read_df <- clean_df_filtered %>%
  filter(taxon_reads > 1)
```




```{r}
## simple bray-curtis (downsampling, VRP)
sample.df <- min_read_df %>%
    #filter(ablg == "e03719") %>% # I think I need to do this sample-by-sample?
    select(sample, taxon, taxon_reads) %>%
    unique() %>% # ensure there are no duplicates because they will cause the pivot to fail later
    group_by(taxon, sample) %>%
    summarise(count = sum(taxon_reads))
  
  ### community-by-species matrix
  # To run the Bray-Curtis and NMDS, we will use the function metaMDS. The function requires a community-by-species matrix.
  
  # reformat the dataframe - wider
comm.wide.df <- sample.df %>%
    pivot_wider(names_from = taxon, values_from = count)
  
  # replace all NAs with 0
comm.wide.df[is.na(comm.wide.df)] <- 0
comm.matrix <- column_to_rownames(comm.wide.df, var = "sample")
  
  # split apart the sample name in the data frame to get the strata
  community.df.sep <- comm.wide.df %>%
    separate(sample, into = c("ablg", "plate", "replicate"), remove = FALSE)
  
  # Calculating relative abundance and creating new dataframe with relative abundance data
  relative.abund <- decostand(comm.matrix, method = "total") # standardization method = total, which probably makes the most sense because it sums over the rows (and not the columns - which would be among replicates... which is what we're curious to calculate the dissimilarity among)
  
  matrix.prop <- as.matrix(relative.abund)
  
  # jaccard similarity (presence/absence)
  jaccard.dist <- vegdist(relative.abund, method = "jaccard")
  
  # generate an easily-filtered dataframe with any comparisons that are > 0.49 dissimilar
  # Creating easy to view matrix
  jaccard_distmat2 <- as.matrix(jaccard.dist, labels = T)
  
  as.data.frame.table(jaccard_distmat2) %>%
  rename(rep1 = Var1, rep2 = Var2, Jaccard_dist = Freq) %>%
  filter(Jaccard_dist > 0.5) 
  
  # Permanova for jaccard 
  set.seed(129)
  jaccard.pool.perm <- adonis2(jaccard.dist~plate, data = community.df.sep, permutations = 999, method = "jaccard")
  
  # Bray-Curtis dissimilarity
  # Calculate distance matrix
  bray.distmat <- vegdist(relative.abund, method = "bray")
  
  # generate an easily-filtered dataframe with any comparisons that are > 0.49 dissimilar
  # Creating a matrix
  bray_distmat2 <- as.matrix(bray.distmat, labels = T)
  # reformat for easier filtering as a table
  bray_df <- as.data.frame.table(bray_distmat2) %>%
    rename(rep1 = Var1, rep2 = Var2, BrayCurtis_dist = Freq)

```
Both Jaccard and Bray-Curtis fare poorly with rare targets, I'd bet.

```{r explore-a-bit-more}
bray_reps <- bray_df %>%
  separate(rep1, into = c("ablg.1", "plate.1", "rep.1")) %>%
  separate(rep2, into = c("ablg.2", "plate.2", "rep.2") ) %>%
  filter(ablg.1 == ablg.2) %>% # only the same ablg number
  filter(plate.1 == plate.2) %>% # compare reps a and b
  filter(rep.1 != rep.2) # but no need to compare replicates with themselves

bray_reps %>%
  ggplot(aes(BrayCurtis_dist)) +
  geom_histogram()

```
The vast majority of a/b replicates are more similar than dissimilar.
258 comparisons; 26 > 0.5

almost exactly 10%


Quickly take a look at any that are more dissimilar:
```{r dissimilar-extraction-replicates}
bray_reps %>%
  filter(BrayCurtis_dist > 0.5) #%>%
 # unique()


```

What about plate replicates?
We would anticipate these being more similar than the extraction replicates, potentially.
```{r plate-replicates-bray}
bray_plates <- bray_df %>%
  separate(rep1, into = c("ablg.1", "plate.1", "rep.1")) %>%
  separate(rep2, into = c("ablg.2", "plate.2", "rep.2") ) %>%
  filter(ablg.1 == ablg.2) %>% # only the same ablg number
  filter(rep.1 == rep.2) %>% # compare plate replicates
  filter(plate.1 != plate.2) # but no need to compare replicates with themselves

bray_plates %>%
  ggplot(aes(BrayCurtis_dist)) +
    geom_histogram()
```

```{r}
bray_plates %>%
  filter(BrayCurtis_dist > 0.5) %>%
  select(ablg.1) %>%
  unique()
```
82/516

```{r}
82/516
```
~16% of PCR replicates were more dissimilar (from 11 samples)

What about summarizing over all replicates. We assume that detections are true detections.
```{r}
presence_df <- min_read_df %>%
  group_by(ablg) %>%
  mutate(presence = ifelse(taxon_reads > 0, "1", "0")) %>%
  select(ablg, alternative_ID, source, taxon, presence) %>%
  unique() %>%
  mutate(detection_method = "DNA")
  
presence_df
```


```{r}
tmp_df <- hardparts_df %>%
  filter(DNASample != "NULL") %>%
  mutate(detection_method = "hardparts") %>%
  rename(taxon = hardparts, alternative_ID = DNASample) %>%
  select(-HardPartsSample, -SpeciesCode, -CommonName, -Notes) %>%
  mutate(source = ifelse(str_detect(alternative_ID, "Cu23"), "scat matrix", NA)) %>%
  mutate(source = ifelse(str_detect(alternative_ID, "AN"), "fecal loop", source)) %>%
  mutate(taxon = str_replace(taxon, " sp.", "")) %>%
  mutate(taxon = str_replace(taxon, "Actinopterygii", "Actinopteri")) %>%
  mutate(taxon = str_replace(taxon, "Stenobrachius leucopsaurus", "Stenobrachius"))


# plot comparison
presence_df %>%
  bind_rows(., tmp_df) %>%
  mutate(taxon = str_replace(taxon, "Stenobrachius leucopsaurus", "Stenobrachius")) %>%
  ggplot(aes(x = alternative_ID, fill = taxon)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(detection_method), cols = vars(source), space = "free", scales = "free") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) 
  
```

```{r}
combo_df <- presence_df %>%
      mutate(taxon = str_replace(taxon, "Leuroglossus", "Leuroglossus schmidti")) %>%
        mutate(taxon = str_replace(taxon, "Sebastes entomelas", "Sebastes")) %>%
    mutate(taxon = str_replace(taxon, "Sebastes zacentrus", "Sebastes")) %>%
  bind_rows(., tmp_df) %>%
    mutate(taxon = str_replace(taxon, "Stenobrachius leucopsaurus", "Stenobrachius")) 

 combo_df$taxon <- sub(" ", "_", combo_df$taxon)

 
 combo_df %>%
  group_by(taxon) %>%
  add_tally() %>%
  filter(source == "scat matrix") %>%
  ggplot(aes(x = reorder(taxon, -n), fill = detection_method)) +
  geom_bar(stat = "count") +
  facet_grid(rows = vars(source), space = "free", scales = "free") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )  +
  labs(x = "taxon",
       y = "number of samples")
  
ggsave("pdf_outputs/NFSfecal_DNAhardparts.pdf", width = 8, height = 6)
```





## Check fecal loops vs. scats


