---
title: "R Notebook"
output: github_document
---

August 2023 - Bogoslof SeaBird Temperature/Pressure casts

Quick look at SeaBird cast data from field collections, August 6-11.


```{r load-libraries}
library(tidyverse)
```
## Notes and to-dos

Need to convert pressure to depth

Need to change UTC to AK time to determine dates/times for each sampling event (1-5).


Eddy field around Bogoslof - Jeremy could pull down the remote sensing data - additional foraging corresponding to eddies

Monthly composite of chlorophyll for August

Argo float - subsurface current field

Ask PMEL if there is any data for this time period

Climate modelers at UW/PMEL Al Herman and Wei, running the models and running the bias correction.



### Confirm that I have all the CTD data onboard.

```{r read-in-all-data}
cumulative_dat <- read_csv("../data/SBE39plus09899_2023-09-12.asc", skip = 11, col_names = F) %>%
  rename(temperature = X1, pressure = X2, date = X3, time = X4) %>%
  separate(date, into = c("day", "month", "year")) %>%
  filter(month != "Jul") # remove test data prior to Bogoslof

cumulative_dat %>%
  group_by(day, month) %>%
  tally() 
```



```{r}
cumulative_dat %>%
 filter(day %in% c(10, 11)) %>%
  ggplot(aes(x = time, y = pressure)) +
  geom_point()


```

That looks like a pretty consistent signal on pressure/depth.



```{r plot-casts-example}
# for one sampling period
cumulative_dat %>%
  filter(day %in% c(10, 11)) %>%
  filter(pressure != 0) %>%
  filter(temperature < 10) %>%
  ggplot(aes(x = temperature, y = -1*pressure)) +
  geom_point() + 
  labs(x = "temperature C",
       y = "pressure dbar*-1")


```

Convert between UTC and AK time by subtracting 8 hours.
