---
title: "Processing data for water samples from 20250221 MiSeq run"
output: html_notebook
---

Processing these data locally for now...
1. Used cutadapt to remove primers
2. unzipped files before going to dada2


```{r load-libraries}
library(dada2)
library(dplyr)

```

```{r}
# file location
path <- "/Users/diana.baetscher/git-repos/NFS-Bogoslof/MiSeq_data/trimmed"

```

```{r}
fnFs <- sort(list.files(path, pattern = "_R1.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern = "_R2.fastq", full.names = TRUE))
# Extract sample names, assuming filenames have format: SAMPLENAME_XXX.fastq
sample.names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
```


```{r}
plotQualityProfile(fnFs[1:2])
plotQualityProfile(fnRs[1:2])
```
Read quality... not amazing universally. Probably good to trim somewhat liberally.



```{r}
# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names
```

Initial filtering truncLen failed miserably. We changed it to be super-stringent to see how that affects downstream processing.
```{r}
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(100,100),
              maxN=0, maxEE=c(2,4), truncQ=2, rm.phix=TRUE,
              compress=TRUE, multithread=TRUE) 
head(out)
```

Those filters worked decently for now - keeping a majority of the reads.

```{r}
# ensure there are no samples without reads
# as these don't work in the next steps
out %>%
  as.data.frame() %>%
  filter(reads.out <1)

# filter the matrix to retain samples with >0 reads
```
### Error rates
```{r}
# fwd error rates
errF <- learnErrors(filtFs, multithread=TRUE)

# reverse error rates
errR <- learnErrors(filtRs, multithread=TRUE)

# plot the erors
p1 <- plotErrors(errF, nominalQ=TRUE)
p2 <- plotErrors(errR, nominalQ=TRUE)

```

### Sample inference

```{r}
# forwards
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)

# reverses
dadaRs <- dada(filtRs, err=errR, multithread=TRUE)

```

Merge paired end reads
```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# Inspect the merger data.frame from the first sample
head(mergers[[1]])
```

Make a sequence table
```{r}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```


```{r}
# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
```
Let's remove the singletons and off-target sequences
```{r}
#seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 168:172]
seqtab2 <- seqtab[,nchar(colnames(seqtab)) %in% 162:182]

```


Remove chimeras
```{r}
seqtab.nochim <- removeBimeraDenovo(seqtab2, method="consensus", multithread=TRUE, verbose=TRUE)
dim(seqtab.nochim)
```
Calculate frequency of chimeras
```{r}
sum(seqtab.nochim)/sum(seqtab2)

```

Track reads through the pipeline
```{r}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))

colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)
```

Again, a lot of data loss at the merge step.



## Export files for taxonomy and samples/ASVs

```{r regseqs-asv-output}
 #make fasta file with ASVs
    asv_seqs=colnames(seqtab.nochim)
    for(i in 1:length(asv_seqs))
    {
        write.table(paste(">ASV",i, sep=""),file="csv_outputs/NFS_MiFish_ASV_seqtab_nochim_20250221_wider.csv", append=TRUE, col.names = F, row.names = F, quote=F)
        write.table(paste(asv_seqs[i], sep=""),file="csv_outputs/NFS_MiFish_ASV_seqtab_nochim_20250221_wider.csv", append=TRUE, col.names = F, row.names = F, quote=F)
    }
```

That's the input for the FASTA blastn search.


Goal: change ASV headers to numbered ASVs that correspond to those output in the FASTA file.
```{r output-asv-table}
# Make map between brief names and full sequences
briefToSeq <- colnames(seqtab.nochim)
names(briefToSeq) <- paste0("ASV", seq(ncol(seqtab.nochim))) # Seq1, Seq2, ...
# Make new sequence table with brief names
st.brief <- seqtab.nochim
colnames(st.brief) <- names(briefToSeq)

# export the pool seq table with brief names:
write.csv(st.brief, file="csv_outputs/NFS_MiFish_ASVtable_20250221_wider.csv")
```

Now move on to the blastn taxonomy search.



## Taxonomy outputs

I ran blastn on Sedna and then taxonkit locally.


```{r taxonomy-files}
library(tidyverse)
taxonomy <- read_tsv("../blast_outputs/blastn_taxlineage_NFS_wider.txt", col_names = F) 

head(taxonomy)

tax_df <- taxonomy %>%
  filter(!str_detect(X16, "environmental")) %>% # get rid of any environmental samples 
  select(X1, X3, X17) %>% #remove unnecessary columns
  group_by(X1, X17) %>% # group by the sequence key and the full taxonomy to reduce duplicate entries
  unique() %>% # doing that reduced the number of entries from 146k to 17k
  filter(!is.na(X17)) %>%
  separate(X17, into=c("kingdom", "phylum", "class", "order", "family", "genus", "species"), sep = ";") %>%
  select(-kingdom) %>%
  dplyr::rename(ASV = X1, perc_id = X3) %>%
  filter(!family %in% c("Hominidae", "Bovidae", "Canidae", "Suidae")) # remove these up front.

  
```

## Taxonomic filtering ##

Using a 1% identity filter!

For species-level matches:
1. Any ASV that matches only a single species.


```{r checking-for-syntax-issues-with-blast-taxonomy-outputs}
tax_asv_top_perc <- tax_df %>%
  mutate(species = ifelse(str_detect(species, " sp\\."), NA, species)) %>% 
  # remove other stuff that isn't on-target or out of range
  filter(!species %in% c("Oncorhynchus kawamurae", "Ammodytes japonicus", "Sebastes baramenuke", "Sebastes cheni", "Sebastes inermis", "uncultured bacterium", "Hexagrammos otakii", "Clupea harengus")) %>% # remove species outside of their range
  filter(family != "Phasianidae") %>%
  group_by(ASV, species) %>%
  slice_max(perc_id, n = 1)
  # take the highest perc_id for a single species (taxonomy) entry per ASV
# this remove entries for a given species/ASV at different percent ids
  
single_spp_taxa <-tax_asv_top_perc %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>%
  left_join(., tax_asv_top_perc, by = "ASV") %>%
  select(-n) %>%
  mutate(taxon = species, taxonomic_level = "species")


```
61 ASVs have a single species taxonomic match.




```{r organize-taxonomy-for-the-rest-of-the-ASVs}
# with the multi-species ASVs, rank the species in the top 1% based on identity
id_ranked_df <- tax_asv_top_perc %>%
  anti_join(., single_spp_taxa, by = "ASV") %>%
  group_by(ASV) %>%
  mutate(id_rank = rank(desc(perc_id), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
  mutate(top_perc = max(perc_id)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%) 
  arrange(ASV) %>%
  mutate(perc_diff = top_perc-perc_id) %>%
  filter(!perc_diff > 1) # 1% difference #### EDIT THIS FOR THE % DIFFERENCE ####
  
```


```{r}
# now determine which ASVs are single-species based on the top 1%
species_level_asv <- id_ranked_df %>%
  group_by(ASV, species) %>%
  tally() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>%
  left_join(., id_ranked_df, by = "ASV") %>%
  select(-n) %>%
  filter(perc_id == top_perc) %>%
  filter(!is.na(species)) %>% # remove the sequences without species-level ids
  mutate(taxon = species, taxonomic_level = "species")
  
```
an additional 23 ASVs that have a single taxonomic identity within the top 1% of blast matches.




```{r non-species-level-matches}
# genus-level
  single_genus <- id_ranked_df %>%
    anti_join(., species_level_asv, by = "ASV") %>% 
    group_by(ASV, genus) %>%
    tally() %>%
    ungroup() %>%
    group_by(ASV) %>%
    tally(name = "n_genera") %>%
    filter(n_genera == 1) %>% # sequences that match only a single genus
    select(-n_genera) %>%
    left_join(., id_ranked_df) %>%
    filter(!is.na(genus)) %>% # get rid of sequences that are NA at the genus level
    mutate(taxonomic_level = "genus") %>%
    mutate(taxon = genus)
  
# single family
single_family <- id_ranked_df %>%
    anti_join(., species_level_asv, by = "ASV") %>% 
    anti_join(., single_genus, by = "ASV") %>%
    group_by(ASV, family) %>%
    tally() %>%
    ungroup() %>%
    group_by(ASV) %>%
    tally(name = "n_families") %>%
    filter(n_families == 1) %>% # sequences that match only a single genus
    select(-n_families) %>%
    left_join(., id_ranked_df) %>%
    filter(!is.na(family)) %>% # get rid of sequences that are NA at the genus level
    mutate(taxonomic_level = "family") %>%
    mutate(taxon = family)

single_family

# single order
  single_order <- id_ranked_df %>%
    anti_join(., species_level_asv, by = "ASV") %>% 
    anti_join(., single_genus, by = "ASV") %>%
    anti_join(., single_family, by = "ASV") %>%
    group_by(ASV, order) %>%
    tally() %>%
    ungroup() %>%
    group_by(ASV) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single order
    select(-n) %>%
    left_join(., id_ranked_df) %>%
    filter(!is.na(order)) %>%
    mutate(taxonomic_level = "order") %>%
    mutate(taxon = order)
  
  # single class
  single_class <- id_ranked_df %>%
    anti_join(., species_level_asv, by = "ASV") %>% 
    anti_join(., single_genus, by = "ASV") %>%
    anti_join(., single_family, by = "ASV") %>%
    anti_join(single_order, by = "ASV") %>%
    group_by(ASV, class) %>%
    tally() %>%
    ungroup() %>%
    group_by(ASV) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single class
    select(-n) %>%
    left_join(., id_ranked_df) %>%
    filter(!is.na(class)) %>%
    mutate(taxonomic_level = "class") %>%
    mutate(taxon = class)
  
  # single phylum
  single_phylum <- id_ranked_df %>%
   anti_join(., species_level_asv, by = "ASV") %>% 
    anti_join(., single_genus, by = "ASV") %>%
    anti_join(., single_family, by = "ASV") %>%
    anti_join(single_order, by = "ASV") %>%
    anti_join(single_class, by = "ASV") %>%
    group_by(ASV, phylum) %>%
    tally() %>%
    ungroup() %>%
    group_by(ASV) %>%
    tally() %>%
    filter(n == 1) %>% # sequences that match only a single phylum
    select(-n) %>%
    left_join(., id_ranked_df) %>%
    filter(!is.na(phylum)) %>%
    mutate(taxonomic_level = "phylum") %>%
    mutate(taxon = phylum)
  
```

```{r combine-taxonomy-dataframes}
# this dataframe still has multiple taxonomic entries per sequence
# only by selecting the taxon column can we fully collapse the df
taxonomy_combined <- bind_rows(single_spp_taxa, species_level_asv, single_genus, single_family, single_order, single_class, single_phylum) %>%
   mutate(top_perc = ifelse(is.na(top_perc), perc_id, top_perc)) %>%
    mutate(id_rank = ifelse(is.na(id_rank), 1, id_rank)) %>%
     mutate(perc_diff = ifelse(is.na(perc_diff), 0, perc_diff))


# how many species-level assignments are <98% identity?
# let's switch those to genus-level IDs
taxonomy_rank_mult_entries <- taxonomy_combined %>%
  select(-perc_diff) %>%
  #filter(id_rank == 1) %>%  # this will make this df smaller, but the assignment is based on the taxon/taxonomic_level, which includes multiple ranks
  mutate(taxon = ifelse(taxonomic_level == "species" & perc_id < 98, genus, taxon)) %>%
  mutate(taxonomic_level = ifelse(taxonomic_level == "species" & perc_id < 98, "genus", taxonomic_level)) 

```


```{r evaluate-taxon-ids-at-any-level}
taxonomy_rank_mult_entries %>%
  select(taxon, taxonomic_level) %>%
  unique()

```
330 unique ASV/taxa
36 unique taxa (excluding ASVs)


## what non-fish are here? 
```{r}
not_Actinopteri <- taxonomy_rank_mult_entries %>%
  ungroup() %>%
  filter(class != "Actinopteri") %>%
  select(species, genus, family, order, class, phylum) %>%
  unique() %>%
  select(species, genus, family, order)

not_Actinopteri
```
all of those make sense.


```{r integrate-taxonomy}
multi_hit_asv_df <- taxonomy_rank_mult_entries %>%
  filter(!is.na(phylum)) %>%
  group_by(ASV) %>%
  add_tally(name = "n_taxonomic_hits") %>%
 # --- STEP 1: extract epithet
  mutate(epithet = word(species, 2)) %>% 

  # --- STEP 2: compute genus- and family-level combined names
  group_by(ASV) %>% 
  mutate(
    # GENUS-LEVEL: "Genus epithet1/epithet2/epithet3"
    combined_genus_epithet = if_else(
      taxonomic_level == "genus" & n_taxonomic_hits < 4,
      paste0(
        first(genus), " ",
        paste(unique(epithet), collapse = "/")
      ),
      taxon
    ),

    # FAMILY-LEVEL: "Genus1/Genus2"
    combined_family_species = if_else(
      taxonomic_level == "family" & n_taxonomic_hits < 4,
      paste(unique(genus), collapse = "/"),
      NA_character_
    )
  ) %>% 
  ungroup() %>% 
  

  # --- STEP 3: choose final hit label
  mutate(
    hit = case_when(
      n_taxonomic_hits == 1 ~ taxon,
      taxonomic_level == "genus"  ~ combined_genus_epithet,
      taxonomic_level == "family" ~ combined_family_species,
      TRUE ~ NA_character_
    )
  )
  

```

```{r add-samples}
samples <- read_csv("csv_outputs/NFS_MiFish_ASVtable_20250221_wider.csv") %>%
  rename(sample = `...1`)


multi_tax_sample_df <- samples %>%
  pivot_longer(2:length(samples), names_to = "ASV", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(., multi_hit_asv_df) %>%
  filter(!is.na(phylum)) %>%
  mutate(hit = ifelse(is.na(hit), taxon, hit)) %>%
  select(sample, reads, hit) %>%
  unique()


multi_tax_sample_df
```

```{r positive-controls}
multi_tax_sample_df %>%
  filter(str_detect(sample, "PC")) %>%
   ggplot(aes(x = sample, y = reads, fill = hit)) +
  geom_bar(stat = "identity", position = "stack") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) 
```

```{r negative-controls}
multi_tax_sample_df %>%
  filter(str_detect(sample, "NC")) %>%
   ggplot(aes(x = sample, y = reads, fill = hit)) +
  geom_bar(stat = "identity", position = "stack") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) 
```

```{r presence-of-sturgeon-in-samples}
multi_tax_sample_df %>%
  filter(str_detect(hit, "Acipenser")) %>%
  arrange(desc(reads))


```
In the samples that we're planning on retaining, how many Sturgeon reads?


```{r organize-sample-df}
# filtering steps by read depth, etc.
multi_tax_sample_df %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  filter(total_reads > 1000) %>%
  mutate(prop = spp_reads/total_reads) %>%
    select(sample, hit, spp_reads, total_reads, prop) %>%
    unique() %>%
  separate(sample, into = c("id", "rep"), remove = F) %>%
  ggplot(aes(x = sample, y = prop, fill = hit)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90)
  ) 
```


```{r}
multi_tax_sample_df %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  select(sample, total_reads) %>%
  #filter(total_reads < 20000) %>%
  mutate(keep = ifelse(total_reads > 1500, "keep", "toss")) %>%
  ggplot(aes(x = log10(total_reads), fill = keep)) +
  geom_histogram()
```

```{r}
multi_tax_sample_df %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  ungroup() %>%
  select(sample, total_reads) %>%
  unique() %>%
  summarise(median(total_reads))

```
The number of reads per replicate is wild. I think I probably end up ditching the low-read depth replicates rather than summing them together.

```{r}
multi_tax_sample_df %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  ungroup() %>%
  select(sample, total_reads) %>%
  unique() %>%
  separate(sample, into = c("id", "rep")) %>%
  ggplot(aes(x = id, y = total_reads, fill = rep)) +
  geom_bar(stat = "identity", position = "stack") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) 
```
Given that amount of variation, let's just see what we have using reasonable filters:

```{r}
samples_to_keep <- multi_tax_sample_df %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  ungroup() %>%
  select(sample, total_reads) %>%
  unique() %>%
  filter(total_reads > 1000) %>%
  # remove controls
  filter(!str_detect(sample, "PC")) %>%
  filter(!str_detect(sample, "NC"))

```
168 sample/replicates with >1000 reads

```{r}
keeper_tax_df <- multi_tax_sample_df %>%
  inner_join(., samples_to_keep) %>%
  group_by(sample, hit) %>%
  mutate(spp_reads = sum(reads)) %>%
  select(sample, hit, spp_reads) %>% # need this here, otherwise the proportions don't sum to 1
  unique() %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>%
  filter(total_reads > 1000) %>%
  select(sample, hit, spp_reads, total_reads) %>%
  unique() %>%
  mutate(prop = spp_reads/total_reads)
```

```{r how-much-sturgeon-in-keepers}
keeper_tax_df %>%
  filter(str_detect(hit, "Acipenser")) %>%
  ggplot(aes(x = prop)) +
  geom_histogram()

keeper_tax_df %>%
  filter(str_detect(hit, "Acipenser")) %>%
  filter(prop > 0.5)

keeper_tax_df %>%
  filter(str_detect(hit, "Acipenser")) %>%
   filter(sample != "e02615-2") %>% # This sample is all sturgeon too.
  ungroup() %>%
  summarise(max(prop)) # filter by read depth or proportion?
# maybe stick with the 56 reads?
```



```{r}
keeper_df <- keeper_tax_df %>%
  filter(sample != "e02615-2") %>% # This sample is all sturgeon too.
  filter(spp_reads > 56 & hit != "Acipenser fulvescens") %>%
  group_by(sample) %>%
  mutate(total_reads = sum(spp_reads)) %>% # recalculate proportions w/o sturgeon
  mutate(prop = spp_reads/total_reads)

# what does that remove?
keeper_tax_df %>%
  filter(sample != "e02615-2") %>% # This sample is all sturgeon too.
  filter(spp_reads < 56)
```


```{r plot-filtered-data}
keeper_df %>%
  ggplot(aes(x = sample, y = prop, fill = hit)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  ) +
  scale_y_continuous(expand = c(0,0))

ggsave("pdf_outputs/eDNA_quicklook_20260120.png", height = 10, width = 15)
```


Next steps:
1. Sum across replicates for a given sample
2. Add in metadata

```{r add-meta-data}
meta <- read_csv("../data/Bogoslof_full_metadata.csv")

```

```{r organize-eDNA-data-to-join}
slim_df <- keeper_df %>%
  ungroup() %>%
  separate(sample, into = c("extraction_ID", "rep"), remove = F) %>%
  select(-sample, -rep, -spp_reads) %>%
  rename(taxa = hit) %>% 
  # take the mean proportion of reads by taxon across replicates
  group_by(extraction_ID, taxa) %>%
  summarise(mean_prop = mean(prop))
```

```{r}
slim_df %>%
  filter(sum(mean_prop) > 1) %>%
  ggplot(aes(x = extraction_ID, y = mean_prop, fill = taxa)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  ) +
  scale_y_continuous(expand = c(0,0))


```

```{r}
# to deal with the mismatch between species comps in different replicates, I needed to do this a bit differently.
# but for species that ARE found in multiple reps, I need to still take the mean of those?
 slim_df <- keeper_df %>%
  ungroup() %>%
  separate(sample, into = c("extraction_ID", "rep"), remove = F) %>%
  rename(taxa = hit) %>% 
  # take the mean proportion of reads by taxon across replicates
  group_by(extraction_ID, taxa) %>%
  summarise(taxa_prop = mean(prop)) %>% # get rid of duplicate entries for spp in multiple replicates
  ungroup() %>%
  group_by(extraction_ID) %>%
  mutate(new_tot_prop = sum(taxa_prop)) %>%
  mutate(new_prop = taxa_prop/new_tot_prop) 
   
  
  
  slim_df %>% 
  ggplot(aes(x = extraction_ID, y = new_prop, fill = taxa)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  ) +
  scale_y_continuous(expand = c(0,0))

 
```




```{r combine}
combined_meta_w_edna <- meta %>%
  left_join(., slim_df) %>%
  rename(proportion_sequencing_reads = new_prop) %>%
  select(-taxa_prop, -new_tot_prop)

```


```{r}
combined_meta_w_edna %>%
  ggplot(aes(x = extraction_ID, y = proportion_sequencing_reads, fill = taxa)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 90)
  ) +
  scale_y_continuous(expand = c(0,0))

ggsave("pdf_outputs/eDNA_by_filter.png", height = 10, width = 15)
```

the other way to provide these data to MML is by station, i.e., summing across the three filters at a given station.

```{r}
combined_meta_w_edna %>%
  write_csv("csv_outputs/Bogoslof_eDNA_w_metadata_20260121.csv")

# single list per sampling location:
combined_meta_w_edna %>%
  select(-proportion_sequencing_reads) %>%
  group_by(Transect, StationKM) %>%
  mutate()

```

